---
title: "Unit 11: In Class Discussion"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
server: shiny
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(cowplot)
library(tidyverse)
library(ggsci)
library(furrr)

plan(multisession, workers = 4)

theme_set(theme_cowplot(font_size = 18))
```


## Course Evaluations

- Available now until May 7th
- Course
- Faculty
    - King
    - Middleton
- [https://evaluation.missouri.edu/](https://evaluation.missouri.edu/)


## Causal inference

> If I see a phenotype after a protein depletion, how should I think more carefully about causation and confounding? I mean depletion itself may also cause other stress in the cell (and it does so!), so I am not sure how to tell if the phenotype is a direct effect of that gene or an indirect downstream effect. Also, if two proteins look functionally connected because loss of one gives similar phenotype as loss of the other, what kind of experiment would help us say this is more likely causation and not just correlation?


## Random Effects Syntax

> This might be a step backwards, but I'm still not confident in random effects in models from last week:

```
lme1 <- lme(WaterTemp ~ AirTempPredicted, data = streams,
            random = (~ 1 | Stream_Name/Site),
            method = "ML")`
```

> Is the above lme formula still a change in the intercept? Is anything actually being divided in Stream_Name/Site to account for nesting, or is this just R syntax?

[Michael Clark's model syntax reference](https://m-clark.github.io/mixed-models-with-R/random_intercepts.html)


## General R model syntax

- `~` = "is modeled by"
- `+` = "and"
- `:` = "by" or "interaction"
- `*` = "main effects and interaction"
- `/` = "nested in"
- `(x | group)` = "separate slope of x per group"
- `I()` = "treat this as a math expression"

Wilkinson, G. N. and C. E. Rogers. 1973. Symbolic description of factorial models for analysis of variance. *Appl. Stat.* 392–399.


## Pseudoreplication

> If there is a level of pseudoreplication in your experiment, is it ok to address the effects you have observed as long as you address the potential issues? Or is this an absolute no-no?


## Quiz 11.3

> The question on the relationship between type 1 and 2 errors/power tripped me up on 11.3. Could we verbally walk through it?


## Power Analysis

> I'd like to know more about when a power analysis should not be done or isn't ideal for a certain project, for instance if all or some data has already been gathered or if sample size ever plays a role in using or not using power analysis?

> Can we go a bit more over when you have to do this analysis and what the consequences are for not doing so? this is not to try and not do anything but to know when this step is ignored in the literature.


## Power {.smaller}

> In a low powered study, how are we supposed to interpret non significant results? How do we know if there’s actually no effect versus just not enough power?

> I understand that you shouldn't go back and retrospectively predict power after collecting data, but what if you weren't the one to design the experiment, but were given a dataset and want to know what your limitations are in analyzing and making conclusions?

> In general, I am a visual learner. I like the histograms to understand type 1 and type 2 errors, but are there any other schematics to illustrate this? I also understand the inverse relationship between Power and Type 2 error but I am having a hard time visualizing it and distinguishing Power from type 1 error.


## Type I and II Error and Power

```{r}
#| panel: sidebar
#| width: 280px
sliderInput(
  "delta",
  "Effect size (d)",
  min = 0.2,
  max = 3.5,
  value = 1.5,
  step = 0.05
)

sliderInput(
  "alpha",
  "Significance level (α)",
  min = 0.01,
  max = 0.20,
  value = 0.05,
  step = 0.01
)

sliderInput(
  "sigma",
  "Std dev (σ)",
  min = 0.5,
  max = 2.5,
  value = 1.0,
  step = 0.05
)

hr()

uiOutput("stats_ui")
```

```{r}
#| panel: fill
plotOutput("distPlot", height = "520px")
```

```{r}
#| context: server
compute <- reactive({
  delta <- input$delta
  alpha <- input$alpha
  sigma <- input$sigma
  mu0 <- 0
  mu1 <- delta

  crit <- mu0 + qnorm(1 - alpha) * sigma
  beta <- pnorm(crit, mean = mu1, sd = sigma)
  power <- 1 - beta
  d <- delta / sigma

  list(
    mu0 = mu0,
    mu1 = mu1,
    sigma = sigma,
    alpha = alpha,
    beta = beta,
    power = power,
    crit = crit,
    d = d
  )
})

output$stats_ui <- renderUI({
  p <- compute()
  tagList(
    tags$table(
      style = "width:100%; font-size:13px; border-collapse:collapse;",
      tags$tr(
        tags$td(style = "padding:4px 0; color:#666;", "α (Type I)"),
        tags$td(
          style = "padding:4px 0; text-align:right; font-weight:600;
                         color:#c04040;",
          sprintf("%.3f", p$alpha)
        )
      ),
      tags$tr(
        tags$td(style = "padding:4px 0; color:#666;", "β (Type II)"),
        tags$td(
          style = "padding:4px 0; text-align:right; font-weight:600;
                         color:#3c5ab0;",
          sprintf("%.3f", p$beta)
        )
      ),
      tags$tr(
        tags$td(style = "padding:4px 0; color:#666;", "Power (1−β)"),
        tags$td(
          style = "padding:4px 0; text-align:right; font-weight:600;
                         color:#28885a;",
          sprintf("%.3f", p$power)
        )
      ),
      tags$tr(
        tags$td(style = "padding:4px 0; color:#666;", "Cohen's d"),
        tags$td(
          style = "padding:4px 0; text-align:right; font-weight:600;",
          sprintf("%.2f", p$d)
        )
      ),
      tags$tr(
        tags$td(style = "padding:4px 0; color:#666;", "Critical value"),
        tags$td(
          style = "padding:4px 0; text-align:right; font-weight:600;",
          sprintf("%.2f", p$crit)
        )
      )
    )
  )
})

output$distPlot <- renderPlot(
  {
    p <- compute()

    x_min <- p$mu0 - 4 * p$sigma
    x_max <- p$mu1 + 4 * p$sigma
    xs <- seq(x_min, x_max, length.out = 600)

    df_h0 <- data.frame(x = xs, y = dnorm(xs, p$mu0, p$sigma), dist = "H0")
    df_h1 <- data.frame(x = xs, y = dnorm(xs, p$mu1, p$sigma), dist = "H1")

    # Shaded region data frames — build as plain x/y to avoid rbind column mismatch
    make_poly <- function(xs_in, ys_in) {
      data.frame(
        x = c(xs_in[1], xs_in, xs_in[length(xs_in)]),
        y = c(0, ys_in, 0)
      )
    }

    # α: tail of H0 beyond crit
    idx_a <- xs >= p$crit
    alpha_df <- make_poly(xs[idx_a], dnorm(xs[idx_a], p$mu0, p$sigma))

    # β: body of H1 below crit
    idx_b <- xs <= p$crit
    beta_df <- make_poly(xs[idx_b], dnorm(xs[idx_b], p$mu1, p$sigma))

    # Power: tail of H1 beyond crit
    idx_p <- xs >= p$crit
    power_df <- make_poly(xs[idx_p], dnorm(xs[idx_p], p$mu1, p$sigma))

    # Label positions
    peak_h0 <- dnorm(p$mu0, p$mu0, p$sigma)
    peak_h1 <- dnorm(p$mu1, p$mu1, p$sigma)
    label_df <- data.frame(
      x = c(p$mu0, p$mu1),
      y = c(peak_h0, peak_h1) + 0.01,
      label = c("H\u2080", "H\u2090"),
      dist = c("H0", "H1")
    )

    ggplot() +
      # Filled regions
      geom_polygon(
        data = alpha_df,
        aes(x, y),
        fill = "#c04040",
        alpha = 0.70
      ) +
      geom_polygon(
        data = beta_df,
        aes(x, y),
        fill = "#3c5ab0",
        alpha = 0.55
      ) +
      geom_polygon(
        data = power_df,
        aes(x, y),
        fill = "#28885a",
        alpha = 0.65
      ) +
      # Distribution curves
      geom_line(
        data = df_h0,
        aes(x, y),
        linewidth = 1.0,
        linetype = "dashed",
        color = "#555555"
      ) +
      geom_line(
        data = df_h1,
        aes(x, y),
        linewidth = 1.2,
        color = "#111111"
      ) +
      # Critical value line
      geom_vline(
        xintercept = p$crit,
        linetype = "dotted",
        linewidth = 0.8,
        color = "#444444"
      ) +
      annotate(
        "text",
        x = p$crit + 0.05 * (x_max - x_min) / 8,
        y = peak_h0 * 0.85,
        label = sprintf("crit = %.2f", p$crit),
        hjust = 0,
        size = 3.5,
        color = "#444444"
      ) +
      # H0 / Ha labels
      geom_text(
        data = label_df,
        aes(x, y, label = label),
        size = 5,
        fontface = "bold"
      ) +
      # Inline region labels
      annotate(
        "text",
        x = p$crit + (x_max - p$crit) * 0.35,
        y = dnorm(p$crit + (x_max - p$crit) * 0.20, p$mu0, p$sigma) *
          0.5,
        label = sprintf("\u03b1 = %.3f", p$alpha),
        color = "#7a1f1f",
        size = 4,
        fontface = "bold"
      ) +
      annotate(
        "text",
        x = p$crit - (p$crit - x_min) * 0.28,
        y = dnorm(p$mu1, p$mu1, p$sigma) * 0.30,
        label = sprintf("\u03b2 = %.3f", p$beta),
        color = "#1a2e70",
        size = 4,
        fontface = "bold"
      ) +
      annotate(
        "text",
        x = p$crit + (x_max - p$crit) * 0.35,
        y = dnorm(p$crit + (x_max - p$crit) * 0.20, p$mu1, p$sigma) *
          0.5,
        label = sprintf("Power = %.3f", p$power),
        color = "#145233",
        size = 4,
        fontface = "bold"
      ) +
      scale_x_continuous(
        name = "Test statistic value",
        breaks = scales::pretty_breaks(8)
      ) +
      scale_y_continuous(
        name = "Density",
        expand = expansion(mult = c(0, 0.08))
      ) +
      theme_minimal(base_size = 14) +
      theme(
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey92"),
        axis.title = element_text(size = 13),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA)
      )
  },
  bg = "white"
)
```


## Experimental design {.smaller}

> In lecture 11.3, when trying to find a sample size with high power, typically you get a larger number. For example, the sample size was 63. A comment was made that this is a large sample size, especially for vertebrates/mammals. Is there a way to account for this or determine the best sample size for vertebrate animals that isn't too large but has good power, especially since factors like money come into effect in real-life situations?

> Is there a way to understand how I should adjust my experimental design (like sample size or replication) if I want to reduce Type II error without inflating Type I error?

> I'd like to go over power analyses more, specifically for a linear model instead of a t-test


## Power analysis by Simulation

- `pwr` is useful, but only to a certain extent
- Simulate data following the model you will fit.
- Use your knowledge of the biology you are studying


## Power by Simulation: OLS model

```{r}
#| echo: true
#| cache: true
#| label: sim_ols
nrep <- 1000
sample_sizes <- c(5, 10, 20, 30)
sds <- c(0.1, 0.5, 1, 2)
slopes <- c(0, 0.25, 0.5, 1, 2, 4)

power_bv <- crossing(n = sample_sizes, sd = sds, slope = slopes) |>
  mutate(
    power = future_pmap_dbl(
      list(n, sd, slope),
      function(n, sd, slope) {
        p_values <- map_dbl(
          .x = 1:nrep,
          .f = function(ii) {
            x <- runif(n, min = 0, max = 1)
            y <- 1 + slope * x + rnorm(n, mean = 0, sd = sd)
            fm <- lm(y ~ x)
            anova(fm)["x", "Pr(>F)"]
          }
        )
        mean(p_values < 0.05)
      },
      .options = furrr_options(seed = TRUE)
    )
  )
```


## Power by Simulation: Bivariate model

```{r}
ggplot(
  power_bv,
  aes(x = factor(n), y = power, color = factor(sd), group = factor(sd))
) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  scale_color_d3(name = "sd") +
  facet_wrap(~slope, labeller = label_both) +
  labs(x = "Sample Size", y = "Power")
```


## What do multilevel models do?

```{r}
ZP <- read_csv("../data/Zooplankton.csv", col_types = "cdd") |>
  mutate(Treatment = factor(Treatment), Block = factor(Block))

ggplot(data = ZP, aes(x = Block, y = Zooplankton, color = Treatment)) +
  geom_point(size = 5) +
  scale_color_bmj()
```


## Two models

"Pooled" model with `lm()` (no Block effect):

```{r}
#| echo: true
fm1 <- lm(Zooplankton ~ Treatment, data = ZP)
```

Multilevel model with `lme()`:

```{r}
#| echo: true
#| message: false
library(nlme)
fm2 <- lme(Zooplankton ~ Treatment, random = ~ 1 | Block, data = ZP)
```


## Model 1

```{r}
#| echo: true
summary(fm1)
```


## Model 1

```{r}
#| echo: true
car::Anova(fm1, type = "III")
```


## What is model 1 really doing?

```{r}
ggplot(ZP, aes(x = Treatment, y = Zooplankton, color = Treatment)) +
  geom_point(
    size = 5,
    position = position_jitter(width = 0.15, seed = 34667)
  ) +
  scale_color_bmj()
```


## Model 2

```{r}
#| echo: true
summary(fm2)
```


## Model 2

```{r}
#| echo: true
car::Anova(fm2, type = "III")
```


## Predictions

```{r}
#| echo: true
grid <- crossing(
  Treatment = levels(ZP$Treatment),
  Block = levels(ZP$Block)
)

fm1_pred <- grid |>
  mutate(Pred_Zoo = predict(fm1, newdata = grid), Model = "lm Model")

fm2_pred <- grid |>
  mutate(Pred_Zoo = predict(fm2, newdata = grid), Model = "lme Model")

Preds <- bind_rows(fm1_pred, fm2_pred)
```


## Predictions

```{r}
ggplot() +
  geom_point(
    data = ZP,
    aes(x = Block, y = Zooplankton, color = Treatment),
    size = 2.5,
    alpha = 0.5
  ) +
  geom_point(
    data = Preds,
    aes(x = Block, y = Pred_Zoo, color = Treatment, shape = Model),
    size = 5
  ) +
  scale_color_bmj() +
  scale_shape_manual(values = c(2, 3))
```


## Power Model 1:

```{r}
#| echo: true
#| cache: true
#| label: sim_lm
grand_mean <- mean(ZP$Zooplankton)
sigma <- sigma(fm1)

nrep <- 1000
n_per_group <- 5
deltas <- seq(0, 1.5, by = 0.1)

power_sim <- future_map_dfr(
  deltas,
  function(delta) {
    group_means <- c(grand_mean, grand_mean - delta, grand_mean - 2 * delta)
    p_values <- map_dbl(
      .x = 1:nrep,
      .f = function(ii) {
        sim_data <- tibble(
          Treatment = factor(rep(
            c("control", "low", "high"),
            each = n_per_group
          )),
          Zooplankton = rnorm(
            n_per_group * 3,
            mean = rep(group_means, each = n_per_group),
            sd = sigma
          )
        )
        fm_sim <- lm(Zooplankton ~ Treatment, data = sim_data)
        anova(fm_sim)["Treatment", "Pr(>F)"]
      }
    )
    tibble(delta = delta, power = mean(p_values < 0.05))
  },
  .options = furrr_options(seed = TRUE)
)
```


## One simulation dataset

```{r}
# Save one example simulated dataset for plotting
example_delta <- 0.5
example_means <- c(
  grand_mean,
  grand_mean - example_delta,
  grand_mean - 2 * example_delta
)
sim_example <- tibble(
  Treatment = factor(rep(c("control", "low", "high"), each = n_per_group)),
  Zooplankton = rnorm(
    n_per_group * 3,
    mean = rep(example_means, each = n_per_group),
    sd = sigma
  )
)

ggplot(sim_example, aes(x = Treatment, y = Zooplankton, color = Treatment)) +
  geom_point(
    size = 5,
    position = position_jitter(width = 0.15, seed = 34667)
  ) +
  scale_color_bmj()
```


## Power Model 1:

```{r}
ggplot(power_sim, aes(x = delta, y = power)) +
  geom_line(color = "firebrick4") +
  geom_point(color = "firebrick4") +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  labs(x = "Between-group Differences", y = "Power for Overall Effect")
```


## Power Model 2:

```{r}
#| echo: true
#| cache: true
#| label: sim_lme
sigma_block <- as.numeric(VarCorr(fm2)["(Intercept)", "StdDev"])
sigma_resid <- sigma(fm2)
n_blocks <- 5

power_sim2 <- future_map_dfr(
  deltas,
  function(delta) {
    group_means <- c(grand_mean, grand_mean - delta, grand_mean - 2 * delta)
    p_values <- map_dbl(
      1:nrep,
      function(ii) {
        block_int <- rnorm(n_blocks, mean = 0, sd = sigma_block)
        sim_data <- crossing(
          Treatment = factor(c("control", "low", "high")),
          Block = factor(1:n_blocks)
        ) |>
          mutate(
            mu = group_means[as.numeric(Treatment)],
            Zooplankton = mu +
              block_int[as.numeric(Block)] +
              rnorm(n(), mean = 0, sd = sigma_resid)
          )
        fm_sim <- lme(
          Zooplankton ~ Treatment,
          random = ~ 1 | Block,
          data = sim_data
        )
        anova(fm_sim)["Treatment", "p-value"]
      }
    )
    tibble(delta = delta, power = mean(p_values < 0.05))
  },
  .options = furrr_options(seed = TRUE)
)
```


## One simulation dataset

```{r}
# Save one example simulated dataset for plotting
example_delta <- 0.5
example_means <- c(
  grand_mean,
  grand_mean - example_delta,
  grand_mean - 2 * example_delta
)
block_int_example <- rnorm(n_blocks, mean = 0, sd = sigma_block)
sim_example2 <- crossing(
  Treatment = factor(c("control", "low", "high")),
  Block = factor(1:n_blocks)
) |>
  mutate(
    mu = example_means[as.numeric(Treatment)],
    Zooplankton = mu +
      block_int_example[as.numeric(Block)] +
      rnorm(n(), mean = 0, sd = sigma_resid)
  )

ggplot(sim_example2, aes(x = Treatment, y = Zooplankton, color = Treatment)) +
  geom_point(
    size = 5,
    position = position_jitter(width = 0.15, seed = 34667)
  ) +
  scale_color_bmj()
```


## Power Model 2:

```{r}
ggplot(power_sim2, aes(x = delta, y = power)) +
  geom_line(color = "firebrick4") +
  geom_point(color = "firebrick4") +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  labs(x = "Between-group Differences", y = "Power for Overall Effect")
```


## Comparison

```{r}
bind_rows(
  power_sim |> mutate(Model = "lm"),
  power_sim2 |> mutate(Model = "lme")
) |>
  ggplot(aes(x = delta, y = power, color = Model)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8) +
  geom_hline(yintercept = 0.05) +
  scale_color_aaas() +
  labs(x = "Between-group Differences", y = "Power for Overall Effect")
```


## Quiz 11.4

> I may have missed this in the lecture, but when looking at questions 2 and 3 on 11-4 quiz, how can you determine if there are false positives in a histogram or datasets when not knowing the FDR?


## Multiple Comparisons {.smaller}

> If my experiment involves a large number of measurements and I analyze the data using ANOVA, does that mean I need a much smaller p-value to claim that my results are statistically significant, or should I use FDR adjusted p-values?

> What are some examples where sequential Bonferroni might be especially important to utilize? Is it worth just going ahead and doing it when doing multiple testing error rate tests?

> To make sure I am understanding the q-value correctly for FDR, the q value helps control for the "additive" nature of multiple testing p-values? It is the acceptance that at least a certain percent of positive results will be false positives but way less than if you just did a bunch of t-tests at alpha = 0.05?


## References
