---
title: "Unit 11: In Class Discussion"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(cowplot)
library(tidyverse)
library(ggsci)
library(furrr)

plan(multisession, workers = 4)

theme_set(theme_cowplot(font_size = 18))
```


## Course Evaluations

- Available now until May 7th
- Course
- Faculty
    - King
    - Middleton
- [https://evaluation.missouri.edu/](https://evaluation.missouri.edu/)

## Quizzes

- 11-3
- 11-4

> I may have missed this in the lecture, but when looking at questions 2 and 3 on 11-4 quiz, how can you determine if there are false positives in a histogram or datasets when not knowing the FDR?


## Random Effects Syntax

> This might be a step backwards, but I'm still not confident in random effects in models from last week:

```
lme1 <- lme(WaterTemp ~ AirTempPredicted, data = streams,
            random = (~ 1 | Stream_Name/Site),
            method = "ML")`
```

> Is the above lme formula still a change in the intercept? Is anything actually being divided in Stream_Name/Site to account for nesting, or is this just R syntax?

[Michael Clark's model syntax reference](https://m-clark.github.io/mixed-models-with-R/random_intercepts.html)


## General R model syntax

- `~` = "is modeled by"
- `+` = "and"
- `:` = "by" or "interaction"
- `*` = "main effects and interaction"
- `/` = "nested in"
- `(x | group)` = "separate slope of x per group"
- `I()` = "treat this as a math expression"

Wilkinson, G. N. and C. E. Rogers. 1973. Symbolic description of factorial models for analysis of variance. *Appl. Stat.* 392–399.


## Pseudoreplication


## Repeatability


## Power Analysis

> I'd like to know more about when a power analysis should not be done or isn't ideal for a certain project, for instance if all or some data has already been gathered or if sample size ever plays a role in using or not using power analysis?

> I'd like to go over power analyses more, specifically for a linear model instead of a t test

> Can we go a bit more over when you have to do this analysis and what the consequences are for not doing so? this is not to try and not do anything but to know when this step is ignored in the literature.


## Power

> In a low powered study, how are we supposed to interpret non significant results? How do we know if there’s actually no effect versus just not enough power?

> I understand that you shouldn't go back and retrospectively predict power after collecting data, but what if you weren't the one to design the experiment, but were given a dataset and want to know what your limitations are in analyzing and making conclusions?


## Experimental design

> In lecture 11.3, when trying to find a sample size with high power, typically you get a larger number. For example, the sample size was 63. A comment was made that this is a large sample size, especially for vertebrates/mammals. Is there a way to account for this or determine the best sample size for vertebrate animals that isn't too large but has good power, especially since factors like money come into effect in real-life situations?


## Causative inference

> If I see a phenotype after a protein depletion, how should I think more carefully about causation and confounding? I mean depletion itself may also cause other stress in the cell (and it does so!), so I am not sure how to tell if the phenotype is a direct effect of that gene or an indirect downstream effect. Also, if two proteins look functionally connected because loss of one gives similar phenotype as loss of the other, what kind of experiment would help us say this is more likely causation and not just correlation?


## Multiple Comparisons


## A menu of MCPs {.smaller}

1. <s>Do nothing</s>
    - Not an option
2. Methods to control the Family-Wise Error Rate (FWER):
    - MCs within a single linear model (e.g. Tukey, etc.; see 08-2)
    - Bonferroni correction
      - Not recommended - overly conservative
    - Sequential Bonferroni procedure
    - Randomization procedures to empirically control FWER
6. Methods to control the False Discovery Rate (FDR)
    - False Discovery Rate Methods
    - _Positive_ False Discovery Rate Methods


## False discovery rate

Proposed by Benjamini and Hochberg [-@Benjamini1995-cw].

- Also see Curran-Everett [-@Curran-Everett2000-qv].

Controls FDR (i.e., rate of Type I errors), rather than FWER

$$\mbox{FDR} = \frac{\mbox{n False Positives}}{\mbox{n All Positives}}$$

e.g., I'm OK with 5% false positives *among the tests I judge as significant*.

Note: False Positive Rate = $\frac{\mbox{n False Positives}}{\mbox{n All Tests}}$

## FDR & FPR

You do 100 correlations. You find 22 are statistically significant at an $\alpha$ = 0.05. Let's say 5 are false positives.

1. What is the FPR?
2. What is the FDR?

$$\mbox{FPR} = \frac{\mbox{n False Positives}}{\mbox{n All Tests}}$$
$$\mbox{FDR} = \frac{\mbox{n False Positives}}{\mbox{n All Positives}}$$


## FDR & FPR

$$FPR = \frac{5}{100} = 0.05$$

$$FDR = \frac{5}{22} = 0.227$$

## Power analysis by Simulation

- `pwr` is useful, but only to a certain extent
- Simulate data following the model you will fit.
- Use your knowledge of the biology you are studying


## Power by Simulation: Bivariate model

```{r}
#| echo: true
#| cache: true
#| label: sim_ols

nrep <- 1000
sample_sizes <- c(5, 10, 20, 30)
sds <- c(0.1, 0.5, 1, 2)
slopes <- c(0, 0.25, 0.5, 1, 2, 4)

power_bv <- crossing(n = sample_sizes, sd = sds, slope = slopes) |>
  mutate(
    power = future_pmap_dbl(
      list(n, sd, slope),
      function(n, sd, slope) {
        p_values <- map_dbl(
          .x = 1:nrep,
          .f = function(ii) {
            x <- runif(n, min = 0, max = 1)
            y <- 1 + slope * x + rnorm(n, mean = 0, sd = sd)
            fm <- lm(y ~ x)
            anova(fm)["x", "Pr(>F)"]
          }
        )
        mean(p_values < 0.05)
      },
      .options = furrr_options(seed = TRUE)
    )
  )
```


## Power by Simulation: Bivariate model

```{r}
ggplot(
  power_bv,
  aes(x = factor(n), y = power, color = factor(sd), group = factor(sd))
) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  scale_color_d3(name = "sd") +
  facet_wrap(~slope, labeller = label_both) +
  labs(x = "Sample Size", y = "Power")
```


## What do multilevel models do?

```{r}
ZP <- read_csv("../data/Zooplankton.csv", col_types = "cdd") |>
  mutate(Treatment = factor(Treatment), Block = factor(Block))

ggplot(data = ZP, aes(x = Block, y = Zooplankton, color = Treatment)) +
  geom_point(size = 5) +
  scale_color_bmj()
```


## Two models

"Pooled" model with `lm()` (no Block effect):

```{r}
#| echo: true

fm1 <- lm(Zooplankton ~ Treatment, data = ZP)
```

Multilevel model with `lme()`:

```{r}
#| echo: true
#| message: false

library(nlme)
fm2 <- lme(Zooplankton ~ Treatment, random = ~ 1 | Block, data = ZP)
```


## Model 1

```{r}
#| echo: true

summary(fm1)
```


## Model 1

```{r}
#| echo: true

car::Anova(fm1, type = "III")
```


## What is model 1 really doing?

```{r}
ggplot(ZP, aes(x = Treatment, y = Zooplankton, color = Treatment)) +
  geom_point(
    size = 5,
    position = position_jitter(width = 0.15, seed = 34667)
  ) +
  scale_color_bmj()
```


## Model 2

```{r}
#| echo: true

summary(fm2)
```


## Model 2

```{r}
#| echo: true

car::Anova(fm2, type = "III")
```


## Predictions

```{r}
#| echo: true

grid <- crossing(
  Treatment = levels(ZP$Treatment),
  Block = levels(ZP$Block)
)

fm1_pred <- grid |>
  mutate(Pred_Zoo = predict(fm1, newdata = grid), Model = "lm Model")

fm2_pred <- grid |>
  mutate(Pred_Zoo = predict(fm2, newdata = grid), Model = "lme Model")

Preds <- bind_rows(fm1_pred, fm2_pred)
```


## Predictions

```{r}
ggplot() +
  geom_point(
    data = ZP,
    aes(x = Block, y = Zooplankton, color = Treatment),
    size = 2.5,
    alpha = 0.5
  ) +
  geom_point(
    data = Preds,
    aes(x = Block, y = Pred_Zoo, color = Treatment, shape = Model),
    size = 5
  ) +
  scale_color_bmj() +
  scale_shape_manual(values = c(2, 3))
```


## Power Model 1:

```{r}
#| echo: true
#| cache: true
#| label: sim_lm

grand_mean <- mean(ZP$Zooplankton)
sigma <- sigma(fm1)

nrep <- 1000
n_per_group <- 5
deltas <- seq(0, 1.5, by = 0.1)

power_sim <- future_map_dfr(
  deltas,
  function(delta) {
    group_means <- c(grand_mean, grand_mean - delta, grand_mean - 2 * delta)
    p_values <- map_dbl(
      .x = 1:nrep,
      .f = function(ii) {
        sim_data <- tibble(
          Treatment = factor(rep(
            c("control", "low", "high"),
            each = n_per_group
          )),
          Zooplankton = rnorm(
            n_per_group * 3,
            mean = rep(group_means, each = n_per_group),
            sd = sigma
          )
        )
        fm_sim <- lm(Zooplankton ~ Treatment, data = sim_data)
        anova(fm_sim)["Treatment", "Pr(>F)"]
      }
    )
    tibble(delta = delta, power = mean(p_values < 0.05))
  },
  .options = furrr_options(seed = TRUE)
)
```


## One simulation dataset

```{r}
# Save one example simulated dataset for plotting
example_delta <- 0.5
example_means <- c(
  grand_mean,
  grand_mean - example_delta,
  grand_mean - 2 * example_delta
)
sim_example <- tibble(
  Treatment = factor(rep(c("control", "low", "high"), each = n_per_group)),
  Zooplankton = rnorm(
    n_per_group * 3,
    mean = rep(example_means, each = n_per_group),
    sd = sigma
  )
)

ggplot(sim_example, aes(x = Treatment, y = Zooplankton, color = Treatment)) +
  geom_point(
    size = 5,
    position = position_jitter(width = 0.15, seed = 34667)
  ) +
  scale_color_bmj()
```


## Power Model 1:

```{r}
ggplot(power_sim, aes(x = delta, y = power)) +
  geom_line(color = "firebrick4") +
  geom_point(color = "firebrick4") +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  labs(x = "Between-group Differences", y = "Power for Overall Effect")
```


## Power Model 2:

```{r}
#| echo: true
#| cache: true
#| label: sim_lme

sigma_block <- as.numeric(VarCorr(fm2)["(Intercept)", "StdDev"])
sigma_resid <- sigma(fm2)
n_blocks <- 5

power_sim2 <- future_map_dfr(
  deltas,
  function(delta) {
    group_means <- c(grand_mean, grand_mean - delta, grand_mean - 2 * delta)
    p_values <- map_dbl(
      1:nrep,
      function(ii) {
        block_int <- rnorm(n_blocks, mean = 0, sd = sigma_block)
        sim_data <- crossing(
          Treatment = factor(c("control", "low", "high")),
          Block = factor(1:n_blocks)
        ) |>
          mutate(
            mu = group_means[as.numeric(Treatment)],
            Zooplankton = mu +
              block_int[as.numeric(Block)] +
              rnorm(n(), mean = 0, sd = sigma_resid)
          )
        fm_sim <- lme(
          Zooplankton ~ Treatment,
          random = ~ 1 | Block,
          data = sim_data
        )
        anova(fm_sim)["Treatment", "p-value"]
      }
    )
    tibble(delta = delta, power = mean(p_values < 0.05))
  },
  .options = furrr_options(seed = TRUE)
)
```


## One simulation dataset

```{r}
# Save one example simulated dataset for plotting
example_delta <- 0.5
example_means <- c(
  grand_mean,
  grand_mean - example_delta,
  grand_mean - 2 * example_delta
)
block_int_example <- rnorm(n_blocks, mean = 0, sd = sigma_block)
sim_example2 <- crossing(
  Treatment = factor(c("control", "low", "high")),
  Block = factor(1:n_blocks)
) |>
  mutate(
    mu = example_means[as.numeric(Treatment)],
    Zooplankton = mu +
      block_int_example[as.numeric(Block)] +
      rnorm(n(), mean = 0, sd = sigma_resid)
  )

ggplot(sim_example2, aes(x = Treatment, y = Zooplankton, color = Treatment)) +
  geom_point(
    size = 5,
    position = position_jitter(width = 0.15, seed = 34667)
  ) +
  scale_color_bmj()

```


## Power Model 2:

```{r}
ggplot(power_sim2, aes(x = delta, y = power)) +
  geom_line(color = "firebrick4") +
  geom_point(color = "firebrick4") +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  labs(x = "Between-group Differences", y = "Power for Overall Effect")
```


## Comparison

```{r}
bind_rows(
  power_sim |> mutate(Model = "lm"),
  power_sim2 |> mutate(Model = "lme")
) |>
  ggplot(aes(x = delta, y = power, color = Model)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8) +
  geom_hline(yintercept = 0.05) +
  scale_color_aaas() +
  labs(x = "Between-group Differences", y = "Power for Overall Effect")
```


## References
