---
title: "Loops_Functions"
format: html
---

## Functions

If you need to do the same thing more than once, consider writing a function.

- Pass some inputs to the function
- Do some stuff within the function
- Get an output

A simple example:

- Inputs: control phenotype, treatment 1 phenotype, treatment 2 phenotype
- Compute the ratio of each treatment to control and calculate the correlation between these
- Return the values of the correlation between all pairs of phenotypes and between the ratios

```{r}

#cc <- function(cp, t1p, t2p)
#{
  
#}

cp <- rnorm(100,10,1)
t1p <- rnorm(100,10,1)
t2p<- rnorm(100,10,1)

r1 <- t1p/cp
r2 <- t2p/cp

cor.out <- tibble(cp_t1p = cor(cp,t1p),
                  cp_t2p = cor(cp, t2p),
                  t1p_t2p = cor(t1p, t2p),
                  r1_r2 = cor(r1,r2))
cor.out


```


## Loops

If you need to iterate through different values or to repeat something many times.

A simple example:

Is the expected correlation between two ratios with the same denominator zero?


```{r}



```


