---
title: "Loops_Functions"
format: html
---

## Functions

If you need to do the same thing more than once, consider writing a function.

- Pass some inputs to the function
- Do some stuff within the function
- Get an output

A simple example:

- Inputs: control phenotype, treatment 1 phenotype, treatment 2 phenotype
- Compute the ratio of each treatment to control and calculate the correlation between these
- Return the values of the correlation between all pairs of phenotypes and between the ratios

```{r}

cc <- function(ii)
{
  cp <- rnorm(100,10,1)
  t1p <- rnorm(100,10,1)
  t2p <- rnorm(100,10,1)
  r1 <- t1p/cp
  r2 <- t2p/cp
  
  cor.out <- tibble(cp_t1p = cor(cp,t1p),
                    cp_t2p = cor(cp, t2p),
                    t1p_t2p = cor(t1p, t2p),
                    r1_r2 = cor(r1,r2))
  return(cor.out)
  
}

cors <- map(.x = 1:1000,
            .f = cc) |> 
  list_rbind()

cors |> 
  pivot_longer(cols = everything()) |> 
  ggplot(aes(value)) +
  geom_histogram() +
  facet_grid(name ~ .)













```


## Loops

If you need to iterate through different values or to repeat something many times.

A simple example:

Is the expected correlation between two ratios with the same denominator zero?


```{r}
iters <- 1000

cors <- numeric(length = iters)
  
for(yy in 1:iters)
{
  cp <- rnorm(100,10,1)
  t1p <- rnorm(100,10,1)
  t2p <- rnorm(100,10,1)
  r1 <- t1p/cp
  r2 <- t2p/cp
  cors[yy] <- cor(r1, r2)
  cat(yy,"\t", cor(r1, r2), "\n")
}

ggplot(tibble(cors), aes(cors)) +
  geom_histogram()

```


